Immune-Related Adverse Events (irAEs) associated with Immune Checkpoint Inhibitor (ICI) Therapy in Bladder Cancer: An Analysis of the FDA Adverse Event Reporting System (FAERS)

### 2025-04-10 code/analysis-code/statistical-analysis.qmd
Updates to Part 4 of MADA Project

>>>>

First, let's fit a logistic regression model to see which factors affect the severity of the reactions:

Let's look at the relationship between severity and sex. We need to recode the sex variable for this:
```{r}
# Create a new binary column: 1 = Serious, 0 = Non-Serious
mydata$serious_bin <- ifelse(mydata$serious == "Serious", 1, 0)
```

Now, we fit the model and look at the plot.

```{r}
library(ggplot2)

logit_model_sex <- glm(serious_bin ~ sex, data = mydata, family = binomial)
summary(logit_model_sex)
mydata$predicted_prob <- predict(logit_model_sex, type = "response")

severity_sex = ggplot(mydata, aes(x = sex, y = predicted_prob)) +
  geom_jitter(width = 0.2, alpha = 0.3, color = "purple") +
  labs(
    title = "Predicted Probability of Serious Event by Sex",
    x = "Sex",
    y = "Predicted Probability"
  ) +
  theme_minimal()

print(severity_sex)

# Define paths
severity_sex_path <- here("results", "figures", "severity_sex.png")
logit_model_sex_path = here("results", "tables", "logit_model_sex.rds")

# Save all four tables to their respective paths
ggsave(filename = severity_sex_path, plot=severity_sex) 
saveRDS(logit_model_sex, file = logit_model_sex_path)

We see that the probability of serious outcomes is highest for sex not specified, followed by males, then females.
There are more cases for males than females, so it is expected that more males report severe cases.
However, it is surprising that even with a low percentage of non-specified sex reports, the severity can be high.
This may indicate underlying issues with the data reporting.
We do see that sex is not a significant predictor of severity.

Let's look at the association of age with severity.
We know that age has a few NA's. Since, this is continuous data, it is better to remove the NA's.

```{r}
mydata_clean <- mydata[!is.na(mydata$patient_age), ]
```

```{r}
logit_model_age <- glm(serious_bin ~ patient_age, data = mydata_clean, family = binomial)
summary(logit_model_age)
mydata_clean$predicted_prob <- predict(logit_model_age, type = "response")

severity_age = ggplot(mydata_clean, aes(x = patient_age, y = predicted_prob)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  labs(
    title = "Predicted Probability of Serious Outcome by Age",
    x = "Patient Age",
    y = "Predicted Probability"
  ) +
  theme_minimal()

print(severity_age)

# Define paths
severity_age_path <- here("results", "figures", "severity_age.png")
logit_model_age_path = here("results", "tables", "logit_model_age.rds")

# Save all four tables to their respective paths
ggsave(filename = severity_age_path, plot=severity_age) 
saveRDS(logit_model_age, file = logit_model_age_path)

```
We see that predicted severity decreases with age.
However, this relationship is also not signficant.

Now, we use machine learning methods (random forest) to assess severity of reaction based on age, sex and drug.
We know that there are NA's in the age.
Let's impute the data (since removing the variables will create a very small dataset).
There were too many NA's for KNN imputation to be used.

```{r}
# Calculate the mean of the 'patient_age' column, excluding missing values
mean_age <- mean(mydata$patient_age, na.rm = TRUE)

# Replace missing values (NA) in 'patient_age' with the calculated mean
mydata$patient_age_imputed <- ifelse(is.na(mydata$patient_age), mean_age, mydata$patient_age)

First, let's split the data:

```{r}
set.seed(123)
library(caret)

train_index <- createDataPartition(mydata$serious_bin, p = 0.7, list = FALSE)
train_data <- mydata[train_index, ]
test_data  <- mydata[-train_index, ]

First, let's ensure that the serious variable is a factor:
```{r}
# Ensure 'serious_bin' and 'sex' is a factor
train_data$serious_bin <- as.factor(train_data$serious_bin)
test_data$serious_bin <- as.factor(test_data$serious_bin)

train_data$sex <- factor(train_data$sex)
test_data$sex <- factor(test_data$sex)
```

First, let's use a basic logistic regression model:
```{r}
# Fit logistic regression model with cross-validation
logit_model <- train(serious_bin ~ patient_age_imputed + sex + suspect_product_active_ingredients, 
                     data = train_data, 
                     method = "glm", 
                     family = "binomial", 
                     trControl = trainControl(method = "cv", number = 5))

Now, let's fit a decision tree (for classification):
```{r}
# Decision Tree (CART) model with cross-validation
dt_model <- train(serious_bin ~ patient_age_imputed + sex + suspect_product_active_ingredients, 
                  data = train_data, 
                  method = "rpart", 
                  trControl = trainControl(method = "cv", number = 5))

Now, let's use a random forest:
```{r}
# Random Forest model with cross-validation
rf_model <- train(serious_bin ~ patient_age_imputed + sex + suspect_product_active_ingredients, 
                  data = train_data, 
                  method = "rf", 
                  trControl = trainControl(method = "cv", number = 5))

Now, let's evaluate all the models:

```{r}
# Accuracy and AUC for Logistic Regression
logit_model$results

# Accuracy and AUC for Decision Tree
dt_model$results

# Accuracy and AUC for Random Forest
rf_model$results

```{r}
library(pROC)
library(ggplot2)
library(here)

# Logistic Regression ROC
logit_roc <- roc(train_data$serious_bin, predict(logit_model, type = "prob")[,2])
logit_roc_plot <- ggroc(logit_roc) + ggtitle("Logistic Regression ROC Curve")
cat("Logistic Regression AUC:", auc(logit_roc), "\n")

# Decision Tree ROC
dt_roc <- roc(train_data$serious_bin, predict(dt_model, type = "prob")[,2])
dt_roc_plot <- ggroc(dt_roc) + ggtitle("Decision Tree ROC Curve")
cat("Decision Tree AUC:", auc(dt_roc), "\n")

# Random Forest ROC
rf_roc <- roc(train_data$serious_bin, predict(rf_model, type = "prob")[,2])
rf_roc_plot <- ggroc(rf_roc) + ggtitle("Random Forest ROC Curve")
cat("Random Forest AUC:", auc(rf_roc), "\n")

# Define file paths
logit_roc_path <- here("results", "figures", "logit_roc.png")
dt_roc_path <- here("results", "figures", "dt_roc.png")
rf_roc_path <- here("results", "figures", "rf_roc.png")

# Save plots
ggsave(filename = logit_roc_path, plot = logit_roc_plot, width = 6, height = 4)
ggsave(filename = dt_roc_path, plot = dt_roc_plot, width = 6, height = 4)
ggsave(filename = rf_roc_path, plot = rf_roc_plot, width = 6, height = 4)

```{r}
cat("\nModel Comparison based on AUC:\n")
cat("Logistic Regression AUC:", auc(logit_roc), "\n")
cat("Decision Tree AUC:", auc(dt_roc), "\n")
cat("Random Forest AUC:", auc(rf_roc), "\n")
```

We see that the Random Forest model has an AUC of 0.806. Thus, this models is able to predict between serious and non-serious cases ~81% of the time.
The logistic regression model is next, able to distinguish ~60% of the time.
The decision tree model performs poorly, possibly due to fitting issues.

Let's fit the random forest model to the test data:
```{r}
# Make predictions on the test data (probabilities for AUC)
rf_pred_prob <- predict(rf_model, newdata = test_data, type = "prob")

# Make predictions on the test data (class labels for evaluation)
rf_pred_class <- predict(rf_model, newdata = test_data, type = "raw")

# ROC curve for Random Forest
rf_roc <- roc(test_data$serious_bin, rf_pred_prob[, 2])  # Using the probability of the 'Serious' class
plot(rf_roc, main = "Random Forest ROC Curve")
cat("Random Forest AUC on Test Data:", auc(rf_roc), "\n")

# Extract the random forest model from the train object
rf_final_model <- rf_model$finalModel

library(randomForest)
# Display feature importance
importance(rf_final_model)

# Plot feature importance
varImpPlot(rf_final_model)

Save the file:
```{r}
# Load library
library(randomForest)

# Define file path
importance_plot_path <- here::here("results", "figures", "rf_feature_importance.png")

# Save the plot
png(filename = importance_plot_path, width = 800, height = 600)
varImpPlot(rf_final_model)
dev.off()

We see that the model does not perform as well on the test data (AUC=0.56), and is able to distinguish the severity only 56% of the time.
The most important feature is the patient age.
However, this variable has imputations and thus may affect the dependability of the model.
Overall, the model performs decently on train data, but not on test data.
More data and variables can help improve the fit.
With this data, this model is not useful in distinguishing between serious and non-serious cases.

Let's create a summary table for all the models:
```{r}
library(pROC)
library(caret)

# Logistic Regression Model
logit_preds <- predict(logit_model, newdata = train_data)
logit_acc <- confusionMatrix(logit_preds, train_data$serious_bin)$overall['Accuracy']
logit_auc <- roc(train_data$serious_bin, predict(logit_model, newdata = train_data, type = "prob")[,2])$auc
logit_sens <- confusionMatrix(logit_preds, train_data$serious_bin)$byClass['Sensitivity']
logit_spec <- confusionMatrix(logit_preds, train_data$serious_bin)$byClass['Specificity']

# Decision Tree Model (now using dt_model)
dt_preds <- predict(dt_model, newdata = train_data)
dt_acc <- confusionMatrix(dt_preds, train_data$serious_bin)$overall['Accuracy']
dt_auc <- roc(train_data$serious_bin, predict(dt_model, newdata = train_data, type = "prob")[,2])$auc
dt_sens <- confusionMatrix(dt_preds, train_data$serious_bin)$byClass['Sensitivity']
dt_spec <- confusionMatrix(dt_preds, train_data$serious_bin)$byClass['Specificity']

# Random Forest Model
rf_preds <- predict(rf_model, newdata = train_data)
rf_acc <- confusionMatrix(rf_preds, train_data$serious_bin)$overall['Accuracy']
rf_auc <- roc(train_data$serious_bin, predict(rf_model, newdata = train_data, type = "prob")[,2])$auc
rf_sens <- confusionMatrix(rf_preds, train_data$serious_bin)$byClass['Sensitivity']
rf_spec <- confusionMatrix(rf_preds, train_data$serious_bin)$byClass['Specificity']

# Create summary data frame with results
model_summary <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(logit_acc, dt_acc, rf_acc),
  AUC = c(logit_auc, dt_auc, rf_auc),
  Sensitivity = c(logit_sens, dt_sens, rf_sens),
  Specificity = c(logit_spec, dt_spec, rf_spec)
)

# Convert to huxtable
model_table <- as_hux(model_summary)

# Format the table
model_table <- set_bold(model_table, 1, 1:ncol(model_table))  # Make the entire header row bold
model_table <- set_align(model_table, 1, 1:ncol(model_table), "center")  # Center-align the entire header row
model_table <- set_all_borders(model_table, 1)

# Print the table
model_table

# Save the table
model_table_path = here("results", "tables", "model_table.rds")
saveRDS(model_table, file = model_table_path)
```
We see that sensitivity is 0 and specificity is high, meaning all the models always predict 'serious'.
Such problems arise due to class imbalances, and may require advanced methods to deal with.

### 2025-04-10 code/eda-code/eda.qmd
Updates to Part 4 of MADA Project

# Convert to huxtable object
huxtable_table <- as_hux(transposed_table)

# Format the table with huxtable (style, borders, etc.)
huxtable_table <- huxtable_table %>%
  set_all_borders(1) %>%
  set_font_size(12) %>%
  set_align("center") %>%
  set_width(1.5)

# Print the huxtable object, this will render correctly in Word
huxtable_table

@@

# Add styling to the table
ht <- ht %>%
    set_bold(1, 1:ncol(ht)) %>%  # Bold the first row (header)
    set_caption("Immune Checkpoint Inhibitors by Organ System and Reaction") %>%
    set_width(1)                  # Set column width to make the table more compact

# Display the table
ht

@@

# Data exploration through figures

>>>>

# Get the current order of attributes
current_order <- transposed_table$Attribute

# Create a new order with subheadings inserted
new_order <- c()
insert_subheading <- function(order_vector, before, subheading) {
  pos <- which(order_vector == before)
  if (length(pos) > 0) {
    c(order_vector[1:(pos-1)], subheading, order_vector[pos:length(order_vector)])
  } else {
    order_vector
  }
}

# Insert subheadings before specific attributes
new_order <- insert_subheading(current_order, "Male", "Sex")
new_order <- insert_subheading(new_order, "Serious", "Severity of reaction")
new_order <- insert_subheading(new_order, "Died", "Outcomes")
new_order <- insert_subheading(new_order, "Healthcare Professional", "Reporter type")
new_order <- insert_subheading(new_order, "US", "Country")

# Create a new table with subheadings
subheading_rows <- data.frame(
  Attribute = setdiff(new_order, current_order),
  stringsAsFactors = FALSE
)
subheading_rows[colnames(transposed_table)[-1]] <- ""

final_table <- bind_rows(
  lapply(new_order, function(x) {
    if (x %in% subheading_rows$Attribute) {
      subheading_rows[subheading_rows$Attribute == x, ]
    } else {
      transposed_table[transposed_table$Attribute == x, ]
    }
  })
)

# Create the final table with title and clean formatting
huxtable_table <- as_hux(final_table, add_colnames = TRUE) %>% 
  # Add a title row
  insert_row(c("Adverse Event Report Summary", rep("", ncol(final_table)-1)), 
             after = 0) %>% 
  # Merge title cells
  merge_cells(1, 1:ncol(final_table)) %>%
  # Basic formatting
  set_all_borders(1) %>%               # All cell borders
  set_align(everywhere, everywhere, "center") %>%  # Center alignment
  set_valign(everywhere, everywhere, "middle") %>% # Vertical center
  set_font_size(10) %>%               # Base font size
  set_font_size(1, everywhere, 12) %>% # Larger font for title
  set_bold(1, everywhere, TRUE) %>%   # Bold title only
  # Ensure no other text is bolded
  set_bold(2:nrow(final_table)+1, everywhere, FALSE)

# Format the title row
huxtable_table <- huxtable_table %>%
  set_background_color(1, everywhere, "gray95") %>%
  set_top_border(1, everywhere, 2) %>%
  set_bottom_border(1, everywhere, 2)

huxtable_table

@@

# FINAL WORKING SOLUTION
ht <- as_hux(organ_system_pivoted, add_colnames = TRUE) %>%
  # Basic formatting
  set_bold(1, everywhere, TRUE) %>%      # Bold headers
  set_align(everywhere, everywhere, "center") %>%
  set_all_borders(brdr(1, "solid")) %>%  # Solid borders
  set_all_padding(4) %>%
  set_font_size(10) %>%
  set_width(0.9)

# Add title as a caption (works in exports)
caption(ht) <- "Immune Checkpoint Inhibitors by Organ System and Reaction"

# Bold specific drug names
drug_cols <- which(colnames(organ_system_pivoted) %in% 
                  c("atezolizumab", "pembrolizumab", "nivolumab", "avelumab"))
if (length(drug_cols) > 0) {
  ht <- ht %>% set_bold(1, drug_cols, TRUE)
}

# Display
ht

@@

```
# Data exploration through figures

### 2025-04-10 products/manuscript/Manuscript.qmd
Updates to Part 4 of MADA Project

To assess whether an ICI has a safety signal associated with a particular irAE, three metrics were used:
1. Proportional reporting ratio (PRR):
This approach compares the proportion of reports for a specific adverse event with a drug of interest to the proportion of the same event reported for all other drugs.

@@

```{r}
#| label: fig-processing1
#| fig-cap: "Study Flowchart"
#| echo: FALSE
knitr::include_graphics(here("assets","Figure1.png"))
```

@@

Firstly, descriptive statistics were used to describe the demographic and other characteristics.
This was followed by creating some descriptive figures.
The PRR, ROR and IC (along with their 95% CI) were then calculated to assess the drug signal.
A random forest model was built using age, sex, and ICI, to distinguish between serious and non-serious cases.

@@

# Display table
ft
```
Pembrolizumab has the highest number of reactions. The largest percentage of reactions are classified as "Other". These include reactions which do not belong to the other organ classes, such as myositis.

@@

A random forest model was built to distinguish between serious and non-serious outcomes using age, sex, and ICI.
Inverse weighting was used to deal with the class imbalance.
The model was built using cross-validation.

@@

# Load and prepare data
rf_metrics <- readRDS("../../results/tables/randomforesttable.rds")

# Create professional flextable
flextable(rf_metrics) |> 
  # Basic formatting
  bold(part = "header") |> 
  border(part = "all", border = fp_border(color = "black", width = 1)) |> 
  # Number formatting
  colformat_double(digits = 3) |>  # Adjust decimal places as needed
  # Alignment
  align(align = "center", part = "header") |> 
  align(align = "right", part = "body", j = which(sapply(rf_metrics, is.numeric))) |> 
  # Layout
  set_table_properties(width = 1, layout = "autofit") |> 
  width(width = c(1.8, rep(1.2, ncol(rf_metrics)-1))) |>  # Adjust column widths
  # Visual enhancements
  bg(part = "header", bg = "#f2f2f2") |>  # Light gray header
  fontsize(size = 10, part = "all") |> 
  padding(padding = 4, part = "all")
```

We see that the accuracy is quite low (~25%). The model is not able to distinguish well between serious and non-serious cases, classifying most non-serious cases as serious. This issue arises due to the class imbalance in the data used to train the model (305 serious cases out of 365).
Advanced machine learning methods may be explored to overcome this problem. Some additional methods will be considered in part 5.

>>>>

To assess whether an ICI has a safety signal associated with a particular irAE, three metrics were used:

1. Proportional reporting ratio (PRR):
This approach compares the proportion of reports for a specific adverse event with a drug of interest to the proportion of the same event reported for all other drugs.

@@

```{r}
#| label: fig-processing1
#| fig-cap: "Study Flowchart"
#| echo: FALSE
knitr::include_graphics(here("assets","Figure1.png"))
```
{{< pagebreak >}}

@@

Firstly, descriptive statistics were used to describe the demographic and other characteristics.
This was followed by creating some descriptive figures.
Two logistic regression models were fit to assess the association by severity with sex and age.
The PRR, ROR and IC (along with their 95% CI) were then calculated to assess the drug signal.
Three different machine learning models were used to distinguish severity on the bases of sex, age, and type of ICI.
Due to missing values, the patient age values were imputed.
First, the data was split into training and testing data (70/30 split). A 5-fold cross-validation was used.

@@

# Display table
ft
```

Pembrolizumab has the highest number of reactions. The largest percentage of reactions are classified as "Other". These include reactions which do not belong to the other organ classes, such as myositis.

@fig-result shows the number of reactions by drug, along with the seriousness of the reaction.

```{r}
#| label: fig-result
#| fig-cap: "Association of severity with sex"
#| echo: FALSE
library(here)
library(knitr)
knitr::include_graphics(here("results","figures","severity_sex.png"))
```

The probability of serious outcomes is highest for sex not specified, followed by males, then females.
There are more cases for males than females, so it is expected that more males report severe cases.
However, it is surprising that even with a low percentage of non-specified sex reports, the severity can be high.
This may indicate underlying issues with the data reporting.
However, sex is not a significant predictor of severity.

@fig-result2 shows the association between severity and age.

```{r}
#| label: fig-result2
#| fig-cap: "Association of severity with age"
#| echo: FALSE
library(here)
library(knitr)
knitr::include_graphics(here("results","figures","severity_age.png"))
```

The predicted severity decreases with age.
However, this relationship is also not signficant.

@@

## Machine Learning Analysis

@tbl--randomforest shows the results of the three machine learning models (logistic regression, decision tree, random forest).

@@

# Load and prepare data
rf_metrics <- readRDS("../../results/tables/model_table.rds")

rf_metrics_cleaned <- rf_metrics[-1, ]  # Removes the second row

flextable(rf_metrics_cleaned) |> 
  bold(part = "header") |> 
  border(part = "all", border = fp_border(color = "black", width = 1)) |> 
  colformat_double(digits = 3) |>  
  align(align = "center", part = "header") |> 
  align(align = "right", part = "body", j = which(sapply(rf_metrics_cleaned, is.numeric))) |> 
  set_table_properties(width = 1, layout = "autofit") |> 
  width(width = c(1.8, rep(1.2, ncol(rf_metrics_cleaned)-1))) |>  
  bg(part = "header", bg = "#f2f2f2") |>  
  fontsize(size = 10, part = "all") |> 
  padding(padding = 4, part = "all")

```

The Random Forest model has an AUC of 0.806. Thus, this models is able to predict between serious and non-serious cases ~81% of the time.
The logistic regression model is next, able to distinguish ~60% of the time.
The decision tree model performs poorly, possibly due to fitting issues.
All the models have the same accuracy. Additionally, sensitivity is 0 and specificity is high, meaning all the models always predict 'serious'.
Such problems arise due to class imbalances, and may require advanced methods to deal with.
The random forest model does not perform as well on the test data (AUC=0.56), and is able to distinguish the severity only 56% of the time.

@fig-result3 shows the feature importance.

```{r}
#| label: fig-result3
#| fig-cap: "Feature importance"
#| echo: FALSE
library(here)
library(knitr)
knitr::include_graphics(here("results","figures","rf_feature_importance.png"))
```

The most important feature is the patient age.
However, this variable has imputations and thus may affect the dependability of the model.
Overall, the model performs decently on train data, but not on test data.
More data and variables can help improve the fit.
With this data, this model is not useful in distinguishing between serious and non-serious cases.
The ROC curves for the different models are available in the Supplementary Material.
